---
title: "Class 9: Halloween Mini-Project"
author: "Emily Hendrickson (PID: A69034780)"
format: pdf
---

**Exploratory Analysis of Halloween Candy**

Loading data 

```{r}
candy_file <- "candy-data.txt"

candy = read.csv(candy_file, row.names=1)
head(candy)
```

**Q1. How many different candy types are in this dataset?**

```{r}
nrow(candy)
```

**Q2. How many fruity candy types are in this dataset?**

```{r}
sum(candy$fruity)
```

**Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Tootsie Pop",]$winpercent
```

**Q4. What is the winpercent value for “Kit Kat”?**

```{r}
candy["Kit Kat",]$winpercent
```

**Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?**

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```

```{r}
library(skimr)
skim(candy)
```

**Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?**

The *winpercent* variable/column is on a different scale. It is values from 0-100%
whereas the other columns are binary (0 or 1).


**Q7. What do you think a zero and one represent for the candy$chocolate column?**

0 and 1 represent True or False. That candy either is not fruity (0) or is fruity (1).


**Q8. Plot a histogram of winpercent values**

```{r}
hist(candy$winpercent)
```

**Q9. Is the distribution of winpercent values symmetrical?**

Yes, you can assume a normal distribution by Shapiro-Wilk test.

```{r}
shapiro.test(candy$winpercent)
```

**Q10. Is the center of the distribution above or below 50%?**

```{r}
sum <- summary(candy$winpercent)
median <- as.numeric(sum["Median"])

```


**Q11. On average is chocolate candy higher or lower ranked than fruit candy?**

Higher

```{r}
choco.winp <- candy$winpercent[as.logical(candy$chocolate)]
fruit.winp <- candy$winpercent[as.logical(candy$fruity)]
mean(choco.winp) > mean(fruit.winp)
```


**Q12. Is this difference statistically significant?**

Yes

```{r}
t.test(choco.winp, fruit.winp)
```

**Q13. What are the five least liked candy types in this set?**

```{r}
head(candy[order(candy$winpercent),], n=5)
```

**Q14. What are the top 5 all time favorite candy types out of this set?**

```{r}
head(candy[order(candy$winpercent, decreasing = TRUE),], n=5)
```

**Q15. Make a first barplot of candy ranking based on winpercent values.**

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col() +
  labs(y = "Candy Types", x = "Win Percent") +
  theme(axis.text.y = element_text(size = 3))

```

**Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?**

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent)) + 
  geom_col() +
  labs(y = "Candy Types", x = "Win Percent") +
  theme(axis.text.y = element_text(size = 3))
```



```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
my_cols[rownames(candy) == "Tootsie Pop"] = "purple"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent)) + 
  geom_col(fill = my_cols) +
  labs(y = "Candy Types", x = "Win Percent") +
  theme(axis.text.y = element_text(size = 3))
```

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "red"
my_cols[rownames(candy) == "Tootsie Pop"] = "purple"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent)) + 
  geom_col(fill = my_cols) +
  labs(y = "Candy Types", x = "Win Percent") +
  theme(axis.text.y = element_text(size = 3))
```


**Q17. What is the worst ranked chocolate candy?**

Sixlets

**Q18. What is the best ranked fruity candy?**

Starburst



**Taking a look at pricepercent**

```{r}
library(ggrepel)

#price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

**Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?**

Twix

**Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?**

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head(candy[ord,c(11,12)], n=5)
```

**Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().**

```{r}
ggplot(candy) + 
  aes(pricepercent, reorder(rownames(candy), pricepercent)) + 
  geom_col() +
  labs(y = "Candy Types", x = "Win Percent") +
  theme(axis.text.y = element_text(size = 3))
```

```{r}
ggplot(candy) + 
  aes(pricepercent, reorder(rownames(candy), pricepercent)) + 
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), xend = 0), col = "gray40") +
  geom_point() +
  labs(y = "Candy Types", x = "Win Percent") +
  theme(axis.text.y = element_text(size = 3))
```

**Exploring correlation structure**

```{r}
library(corrplot)
```

```{r}
cij <- cor(candy)
corrplot(cij)
```

**Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?**

Fruity and chocolate are anti-correlated and have a negative value.

**Q23. Similarly, what two variables are most positively correlated?**

winpercent and chocolate are the most postively correlated.

**PCA**

```{r}
pca <- prcomp(candy, scale = T)
summary(pca)
```


```{r}
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2")

```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) +
        aes(PC1, PC2, size=winpercent/100, text=rownames(my_data), 
        label = rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
#library(plotly)
#ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot( pca$rotation[,1], las=2, ylab = "PC1 Contribution")
```

**Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?**

Fruity, hard, and pluribus. These make sense because most halloween candy is either chocolatey or fruity, most hard candy is fruity, and many pluribus candy (like skittles) are also fruity. 
